---
title: "Checking generalized additive models"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 9,
  fig.height = 4.5,
  out.width = "100%",
  cache = TRUE,      # cache chunk results so re-deploys skip the slow fits
  autodep = TRUE     # invalidate a cached chunk when an upstream chunk changes
)
```

```{r setup, message = FALSE}
library(dScoreTest)
library(mgcv)
```

This article walks through `gof_test()` and `compare_models()` on
`mgcv::gam` fits. Because these examples lean on `grf` and refit GAMs across
sample splits, they are slower than the package vignette --- this is a
web-only article (not run by `R CMD check`).

## Simulation example

We simulate from the four-term additive truth in `mgcv::gamSim(eg = 1)`,
$$y = f_0(x_0) + f_1(x_1) + f_2(x_2) + f_3(x_3) + \varepsilon,$$
where $f_3 \equiv 0$ and $f_0, f_1, f_2$ are nonlinear. The covariates are
independent uniforms.

```{r data}
set.seed(42)
dat <- gamSim(eg = 1, n = 400, dist = "normal", scale = 2, verbose = FALSE)
```

### Goodness of fit

`gof_test()` checks the *functional form* of a fitted model against a
nonparametric alternative built with a regression forest. A correctly
specified additive smooth model is not rejected:

```{r gof-ok}
fit.smooth <- gam(y ~ s(x0) + s(x1) + s(x2) + s(x3), data = dat)
gof_test(fit.smooth)
```

Formally, the above tests that the regression function $E[Y \mid X_0,X_1,X_2,X_3]$ is an additive function of $X_0,X_1,X_2,X_3$. 
In contrast, if we specify that the regression function is a *linear function* of $X_0,X_1,X_2,X_3$, the model is rejected. 

```{r gof-mis}
fit.linear <- lm(y ~ x0 + x1 + x2 + x3, data = dat)
gof_test(fit.linear)
```

In fact, since $f_3=0$ in the data-generating mechanism, an additive regression model without  `s(x3)` is still well-specified. 

```{r gof-drop-null}
fit.drop3 <- gam(y ~ s(x0) + s(x1) + s(x2), data = dat)
gof_test(fit.drop3)   # f3 = 0, still well-specified given (x0, x1, x2)
```

**Note**: `gof_test` should **not** be used for testing significance of a predictor, as illustrated below.

```{r}
fit.drop23 <- gam(y ~ s(x0) + s(x1), data=dat)
gof_test(fit.drop23)
```

Since the model is only fitted on `x0` and `x1`, the above tests 
$$ E[Y \mid X_0, X_1] = f_0(x_0) + f_1(x_1) $$
as opposed to 
$$ E[Y \mid X_0, X_1, X_2, X_3] = f_0(x_0) + f_1(x_1), $$
simply because `gof_test` does not see predictors `x_1` and `x_2` which do not appear in the model's formula (you can see this from the printout `y ~ X, with X consists of x0, x1`).
To ask whether one or more *omitted* covariates carries signal, use `compare_models()`, which gives the hunt access to the alternative's covariates.

### Model comparison

`compare_models()` tests a null model against a larger alternative model and
detects signal living in the alternative's extra terms. Similar to ANOVA, 
this can be used to test significance of one or more predictors. For example, 
we can test the significance of $f_2$ by fitting a model without `s(x2)` and 
comparing it against the full model. 

```{r cmp-mis}
fit.full <- gam(y ~ s(x0) + s(x1) + s(x2) + s(x3), data = dat)
res <- compare_models(fit.drop23, fit.full)
res
```

This p-value correctly measures the significance of predictors `x2, x3`. 

### Diagnostics

`plot()` on a `dScoreTest` object shows diagnostic panels. For a rejected test, the hunted direction `h` lines up with the residuals, so `L = resids * h` has a clearly non-zero mean. Indeed, we see a large positive slope in the 'resids vs hunt' plot below.

```{r plot}
plot(res)
```

`summary()` adds a raw-vs-debiased comparison of the statistic, so you can
see how much the outer orthogonalization step moved it:

```{r summary}
summary(res)
```

### Hunt styles

All three hunts are available for GAMs. The optimal hunt is the default; the WLS hunt is simpler and sometimes nearly as powerful:

```{r styles}
compare_models(fit.drop23, fit.full, hunt.style = "optimal")
compare_models(fit.drop23, fit.full, hunt.style = "wls")
```

## Example with prostate cancer data 

In this section, we follow Wakefield (*Bayesian and Frequentist Regression
Methods*, §12) to fit GAMs for analyzing the prostate cancer data in 
Stamey et al. (1989) and Tibshirani (1996). 

```{r prostate-data}
prostate <- read.table(
  "https://hastie.su.domains/ElemStatLearn/datasets/prostate.data",
  header = TRUE)
prostate$train <- NULL
colnames(prostate)[c(1, 2, 4, 6, 9)] <-
  c("log.can.vol", "log.weight", "log.BPH", "log.cap.pen", "log.PSA")
head(prostate)
```

We want to use data to model the continuous outcome variable `log.PSA`, the log
of prostate specific antigen (PSA), which is a biomarker to predict the
clinical stage of cancer. 

```{r prostate-data2}
with(prostate, hist(log.PSA, breaks=15))
```

The dataset also measures the following covariates:

- `log.can.vol`: log of cancer volume (cc), from the digitized tumor area
  times a thickness.
- `log.weight`: log of the prostate weight (grams).
- `age`: age of the patient (years).
- `log.BPH`: log amount of benign prostatic hyperplasia, a noncancerous
  enlargement of the gland, as an area (cm²).
- `svi`: seminal vesicle invasion, a 0/1 indicator of whether cancer cells
  have invaded the seminal vesicle.
- `log.cap.pen`: log capsular penetration, the linear extent (cm) of cancer
  extending into the prostate capsule.
- `gleason`: the Gleason score, a measure of tumor aggressiveness. Each of the
  two largest cancer areas is graded 1--5 (5 most aggressive) and the two
  grades are summed.
- `pgg45`: the percentage of Gleason scores that are 4 or 5.

In particular, we want to model and visualize how `log.PSA` varies according 
to `log.can.vol` and `log.weight`, while accounting for other covariates. We 
fit a GAM model with penalized regression splines on each covariate, except for 
factors `svi` and `gleason`, which are introduced as simple linear terms. 
In particular, we fit two models for the joint effect of log cancer volume and log weight:
one **additive** (no interaction) and one that adds a **tensor-product interaction** between them. 
Writing the bivariate part as `ti()` main effects plus a `ti()`
interaction is the functional-ANOVA decomposition, so the interaction term
captures *only* the interaction.

```{r prostate-fit}
fit.0 <- gam(log.PSA ~ ti(log.can.vol, k = 6) + ti(log.weight, k = 6) +
                       s(age,         k = 7, bs = "cr") +
                       s(log.BPH,     k = 7, bs = "cr") +
                       s(log.cap.pen, k = 7, bs = "cr") +
                       s(pgg45,       k = 7, bs = "cr") +
                       svi + gleason,
             data = prostate, method = "GCV.Cp")

fit.1 <- gam(log.PSA ~ ti(log.can.vol, k = 6) + ti(log.weight, k = 6) +
                       ti(log.can.vol, log.weight, k = c(6, 6)) +   # interaction
                       s(age,         k = 7, bs = "cr") +
                       s(log.BPH,     k = 7, bs = "cr") +
                       s(log.cap.pen, k = 7, bs = "cr") +
                       s(pgg45,       k = 7, bs = "cr") +
                       svi + gleason,
             data = prostate, method = "GCV.Cp")
```

The two fitted surfaces are visualized as follows.

```{r prostate-surfaces, fig.width = 10, fig.height = 5}
par(mfrow = c(1, 2))
vis.gam(fit.0, view = c("log.weight", "log.can.vol"), theta = 35, phi = 25,
        zlab = "fitted log.PSA", main = "without interaction")
vis.gam(fit.1, view = c("log.weight", "log.can.vol"), theta = 35, phi = 25,
        zlab = "fitted log.PSA", main = "with interaction")
```

### Is the interaction model adequate?

Before testing the interaction, check that the model with interaction is itself well-specified:

```{r prostate-gof, fig.width = 9, fig.height = 4.5}
set.seed(42)
gof.1 <- gof_test(fit.1)
gof.1
plot(gof.1)
```

There is no indication of misspecification.

### Testing for the interaction

`compare_models()` tests the additive null `fit.0` against the
interaction alternative `fit.1`:

```{r prostate-compare}
set.seed(42)
compare_models(fit.0, fit.1)
```

For comparison, `mgcv`'s own approximate significance test for the
interaction term `ti(log.can.vol,log.weight)` is borderline significant. 

```{r prostate-mgcv}
summary(fit.1)$s.table
```


### Stabilizing across sample splits

With only `r nrow(prostate)` observations, a single debiased score test is
sensitive to the random hunt/test split. Re-running it gives a spread of
p-values:

```{r prostate-replicate}
set.seed(42)
replicate(10, compare_models(fit.0, fit.1)$p.val)
```

A practical remedy is to aggregate over many splits with a heavy-tailed 
combination test, which is robust to the
strong dependence between splits (cf. the harmonic-mean p-value of Wilson, 2019):

```{r prostate-hmp}
set.seed(42)
pvals <- replicate(200, compare_models(fit.0, fit.1)$p.val)
1 / mean(1 / pvals)   # harmonic-mean p-value
```

The aggregated p-value seems to agree with the mgcv's result. 
