---
title: "Link function in GLM"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 9,
  fig.height = 4.5,
  out.width = "100%",
  cache = TRUE,      # cache chunk results so re-deploys skip the slow fits
  autodep = TRUE     # invalidate a cached chunk when an upstream chunk changes
)
```

```{r setup, message = FALSE}
library(dScoreTest)
```

The package can be used to check and compare (generalized) linear models with a non-canonical link function. Based on covariates $X_1, X_2, X_3$, suppose the response is generated as $$ Y = 5 \exp(X_1 + X_3) + \varepsilon, \quad \mathbb{E}[\varepsilon \mid X] = 0. $$ In other words, the regression model is specified as $$ \log \mathbb{E}[Y \mid X] = \alpha + \beta_1 X_1 + \beta_2 X_2 + \beta_3 X_3.$$ To fit this generalized linear model, we must use the *non-canonical link* `family = gaussian(link = "log")`. Here is a simple simulation.

```{r}
set.seed(42)
n <- 1000
X <- matrix(rnorm(n * 3), nrow = n)
X[,2] <- X[,2] + 0.2 * X[,1] - 0.3 * X[,3] 
X[,3] <- X[,3] + 0.1 * X[,1]
f <- 5 * exp(X[,1] + X[,3]) 
y0 <- f + 10 * rnorm(n)
plot(X[,1] + X[,3], y0, pch=20, cex=0.5)
curve(5 * exp(x), -5, 5, add=TRUE, lwd=2, col="red")
```

For testing the model specification, we use:
```{r}
# use a start value to help fitting
dat <- data.frame(y0=y0, X1=X[,1], X2=X[,2], X3=X[,3])
fit.0 <- glm(y0 ~ ., family = gaussian(link = "log"), data=dat, start=rep(1,4))
fit.0
test.0 <- gof_test(fit.0)
test.0
```

Incorrect specification of the link can be detected by the goodness-of-fit test.
```{r}
# identity link
fit.1 <- glm(y0 ~ ., family = gaussian(), data=dat, start=rep(1,4))
test.1 <- gof_test(fit.1)
test.1
plot(test.1)
```

We can further use `compare_models` to assess the significance of predictors:
```{r}
# significance of X2
fit.drop2 <- glm(y0 ~ X1 + X3, family = gaussian(link = "log"), 
                  data=dat, start=rep(1,3))
compare_models(fit.drop2, fit.0)
# compare with anova
anova(fit.0, fit.drop2)

# significance of (X2, X3) together
fit.drop23 <- glm(y0 ~ X1, family = gaussian(link = "log"), 
                  data=dat, start=rep(1,2))
compare_models(fit.drop23, fit.0)
# compare with anova
anova(fit.0, fit.drop23)
```

